---
title: "dataprep: data preprocessing and plots"
output: rmarkdown::html_vignette
vignette: >
  %\documentclass{article}
  %\VignetteIndexEntry{dataprep: data preprocessing and plots}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dataprep)
library(ggplot2)
library(scales)
```
## Figure 1. Line plots for variables with names that are essentially numeric and logarithmic
```{r fig.height=4, fig.show='hold', fig.width=6}
# Descriptive statistics
descplot(data,5,65)
```

## Figure 2. Line plots for variables whose names are essentially numeric and logarithmic
```{r fig.height=3, fig.show='hold', fig.width=4}
# Selected descriptive statistics, equal to descdata(data,5,65,c('na','min','max','IQR'))
descplot(data,5,65,c(2,7:9))
```

## Figure 3. Bar charts for the type of variable names that is character
```{r fig.height=5, fig.show='hold', fig.width=6}
# Descriptive statistics
descplot(data1,3,7)+
  ggplot2::theme(axis.text.x=ggplot2::element_text(angle=30,hjust=1,vjust=1.1))
```

## Figure 4. Bar charts for the type of variable names that is character
```{r fig.height=2.5, fig.show='hold', fig.width=6}
# Selected descriptive statistics, equal to descplot(data1,3,7,c('min','max','IQR'))
descplot(data1,3,7,7:9)+
  ggplot2::theme(axis.text.x=ggplot2::element_text(angle=30,hjust=1,vjust=1.1))
```

## Figure 5. Particle number size distributions in logarithmic scales
```{r fig.height=4, fig.show='hold', fig.width=6}
# Top and bottom percentiles
percplot(data,5,65,4)
```

## Figure 6. Particle number size distributions in logarithmic scales with only one part
```{r fig.height=2.5, fig.show='hold', fig.width=6}
# Top percentiles
percplot(data,5,65,4,part=1)
```

## Figure 7. Particle number size distributions in logarithmic scales with only one part
```{r fig.height=2.5, fig.show='hold', fig.width=6}
# Bottom percentiles
percplot(data,5,65,4,part=0)
```

## Figure 8. Percentiles of modes in linear scales
```{r fig.height=5, fig.show='hold', fig.width=6}
# Top and bottom percentiles
percplot(data1,3,7,2)+
  ggplot2::theme(axis.text.x=ggplot2::element_text(angle=30,hjust=1,vjust=1.1))
```

## Figure 9. Percentiles of modes in linear scales with only one part
```{r fig.height=3, fig.show='hold', fig.width=6}
# Top percentiles
percplot(data1,3,7,2,part=1)+
  ggplot2::theme(axis.text.x=ggplot2::element_text(angle=30,hjust=1,vjust=1.1))
```

## Figure 10. Percentiles of modes in linear scales with only one part
```{r fig.height=3, fig.show='hold', fig.width=6}
# Bottom percentiles
percplot(data1,3,7,2,part=0)+
  ggplot2::theme(axis.text.x=ggplot2::element_text(angle=30,hjust=1,vjust=1.1))
```


